Estimating duration using survival analysis

### 2023-06-09 cis-perfect-testing.tex
Fix cross-file referencing

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
\todo{Ref ATACCC chapter and in following paragraph}
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{imperf-test}).
\todo{Ref chapter incl misclassification}
Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.

@@

The reason for this approach, over modelling biomarkers (as in \cref{ATACCC}) is due to two factors.
\todo{Ref ATACCC chapter}
First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.

The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
False negatives are not dealt with and left to \cref{imperf-test}.
\todo{Ref chapter incl misclassification}

@@

Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{intro:sec:cis}).
\todo{ref to description of CIS}
Consider that, based on the estimates in \cref{ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
That is the subject of \cref{imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.
\todo{Ref chapter incl misclassification}

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{imperf-test}).

@@

I have also considered a model including individuals without any observed episodes (see \cref{appendix:total-model} for details).
These individuals may have truncated episodes or may not have been infected at all.
However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).

@@

In this section, I consider the various parameterisation of the survival function that are available.
This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{ATACCC}.
\todo{Ref ATACCC chapter}

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{ATACCC} and the meta-analysis of \textcite{cevikShedding}.
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{ATACCC} analysis.
\todo{Ref ATACCC chapter}
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The analysis in \cref{ATACCC} can be used to provide prior information for the CIS analysis.
\todo{Ref ATACCC chapter here and next few}
In particular, the ATACCC study frequently samples individuals early in the infection.
Therefore, the survival distribution from that analysis should be reliable through this period.
CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.

The model structure assumed in \cref{ATACCC} leads to positive correlation in the posterior estimates of the hazard.
It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autocite{goudieJoining}).
Then, use a Beta process to add additional uncertainty to this.

There are two reasons to increase the uncertainty relative to the \cref{ATACCC} posterior.
First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
Second, the results may not fully generalise between the two studies.
The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{intro:sec:data}).

The first step, approximating the \cref{ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).

@@

I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{ATACCC} with an inflated tail to represent what is seen within CIS.
\todo{Ref ATACCC chapter}
The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.

@@

\appendix

\chapter{Including all CIS individuals in survival analysis}\label{perf-test:sec:total-model} \label{appendix:total-model}

\todo[inline]{Add an intro to this and maybe a plot of results in simulation}

We now consider all $N$ individuals in the CIS cohort who have at
least one test during the period of interest, this includes those
without an observed infection, and assume that each is equally likely to
be infected. As before, $i = 1, \dots, n_a$ indexes episodes but now
$j = 1, \dots, N$ indexes individuals. All episodes have an associated
individual, each individual may have any number (including 0) episodes.
Admissible regions are only defined for episodes, and have not changed
from the previous section. The truncated region is defined for each
individual, in addition we consider the combined truncated region which
is episodes that would be truncated without conditioning on which
individual the episode occurs in:
$\Omega^C = \{ (b, e, j) : (b, e) \in \Omega_j^C \}$.

Denote by $n_\text{tot}$ the total number of episodes across all $N$
individuals, regardless of whether they are admissible or truncated (we
know there are no inadmissible episodes by definition).

For any infection, the probability of it being admissible for episode
$i$ is $\frac{1}{N} p_{ia}$. That is, the probability that the
infection occurs in the individual corresponding to the individual in
which episode $i$ ($1/N$ by the assumption that episodes are equally
likely to occur in any of the $N$ individuals) multiplied by the
probability that the episode is admissible for episode $i$ conditional
on it occurring in the relevant individual ($p_{ia}$).

The total number of truncated infections that occur is
$n_t = n_\text{tot} - n_a$. Conditional on an infection occurring in
individual $j$, its probability of being truncated is $p_{jt}$ (as
defined previously). Therefore, the overall probability of an episode
being truncated is $p_t = \frac{1}{N} \sum_{j=1}^N p_{jt}$.

Consider our data as there being one admissible episode for each
observed episode, and all other episodes being truncated. As in the
previous section, conditional on $n_\text{tot}$, we have a multinomial
likelihood.
\begin{align}
p(\text{data} \mid \theta)
&= \sum_{n_\text{tot}=n_a}^\infty p(\text{data} \mid n_\text{tot}, \theta) p(n_\text{tot}) \\
&= \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) \left( \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \right) \\
\end{align}

In this case, it would be computationally feasible to augment the data
with $n_\text{tot}$ directly (as it is only a single parameter).
However, implementation is easier if we can use a closed-form for the
sum (as in the previous section) as it allows the use of standard
implementations of NUTS such as Stan, which cannot handle discrete
parameters.

Assuming $N_\text{tot} \dist \NegBin(\mu, r)$:
\begin{align}
&\sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} p(n_\text{tot} \mid \lambda) p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \frac{\lambda^n_\text{tot} e^{-\lambda}}{n_\text{tot}!} p(\lambda) d\lambda \\
&= \int \sum_{n_\text{tot}=n_a}^\infty \frac{1}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}-n_a} \lambda^{n_\text{tot}-n_a} \lambda^{n_a} e^{-\lambda} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) \sum_{n_t=0}^\infty \frac{1}{n_t!} p_t^{n_t} \lambda^{n_t} d\lambda &n_t = n-n_a\\
&= \int \lambda^{n_a} e^{-\lambda} p(\lambda) e^{\lambda p_t} d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} p(\lambda) d\lambda \\
&= \int \lambda^{n_a} e^{-\lambda(1 - p_t)} \frac{b^a}{\Gamma(a)} \lambda^{a-1} e^{-b\lambda} \lambda d\lambda \\
&= \int \frac{b^a}{\Gamma(a)} \lambda^{a+n_a-1} e^{-(b+1-p_t)\lambda} \lambda d\lambda \\
&= \frac{b^a}{\Gamma(a)} \frac{\Gamma(a+n_a)}{(b+1-p_t)^{a+n_a}} \\
&\propto (b+1-p_t)^{-(a+n_a)} \\
&= (r/\mu + 1 - p_t)^{-(r+n_a)} \\
&\propto(r + \mu (1- p_t))^{-(r+n_a)}
\end{align}

Which gives the full posterior density as:
\begin{align}
p(\theta \mid \text{data})
&\propto \frac{p(\theta) \prod_{i=1}^{n_a} p_{ia}}{(r + \mu(1 - p_t))^{r+n_a}}
\end{align}

As before, being able to reconstruct the posterior of $n_\text{tot}$
using its full conditional is useful.
\begin{align}
p(n_\text{tot} \mid \text{data}, \theta)
&\propto p(\text{data} \mid \theta, n_\text{tot}) p(n_\text{tot}) \\
&= \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} \left( \prod_{i=1}^{n_a} \frac{1}{N} p_{ia} \right) p_t^{n_\text{tot}-n_a} p(n_\text{tot}) \\
&\propto \frac{n_\text{tot}!}{(n_\text{tot}-n_a)!} p_t^{n_\text{tot}} \frac{\Gamma(r + n_\text{tot})}{n_\text{tot}!} \left( \frac{\mu}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma(r + n_\text{tot})}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}}  \\
&\propto \frac{\Gamma((r + n_a) + (n_\text{tot}- n_a))}{(n_\text{tot}-n_a)!} \left( \frac{\mu p_t}{r + \mu} \right)^{n_\text{tot}-n_a}  \\
\end{align}
which is a negative binomial pmf with size parameter $r + n_a$ and
probability parameter $\frac{r+\mu(1-p_t)}{r+\mu}$.

\listoftodos

\end{document}

>>>>

The estimation of incidence from prevalence is sensitive to the tail of the duration distribution.
In \cref{E-ATACCC} I estimated the duration distribution from the ATACCC data, however, this study suffered from a small sample size and only 20 days of follow-up.
Therefore, the estimation of the tail of the duration distribution is driven by model assumptions.
This chapter develops methods to estimate the distribution using CIS data, which has less frequent sampling but a large sample size and long follow-up.

I take a survival analysis approach to model the coarse sampling of the CIS (see \cref{perf-test:sec:problem}).
Flat priors can be unintentionally very informative, and hence weakly informative priors should be used; in addition, the prior can be used to combine the information in the CIS with the results of \cref{E-ATACCC} (see \cref{perf-test:sec:parameters-priors}).
Through a simulation study (see \cref{perf-test:sec:simulation-study}), I show that these methods perform well, and in this context are not sensitive to model or prior specification.
However, application in reality must account for misclassification bias, the subject of \cref{E-imperf-test}.

@@

I define an individual as \emph{detectable} if they would test positive if tested, in the absence of misclassification error (which is dealt with in \cref{E-imperf-test}).
Episode $i$'s \emph{duration of positivity}, $D_i$ is the number of days that they are detectable for; I use time discretised to days, therefore, the duration is at least 1 day.
Note that the relationship between positivity (\ie the result of a PCR test) and infectiousness (\ie an individual's ability to pass on the virus) is complex, \todo{ref for PCR pos vs infectiousness} and beyond the scope of this thesis.
I define an episode as starting at time $B_i$ and ends at $E_i$, hence $D_i = E_i - B_i + 1$.

@@

The reason for this approach, over modelling biomarkers (as in \cref{E-ATACCC}) is due to two factors.
First, the large sample size in CIS means that approaches where the number of parameters scales with the number of observed infections (as it does with random effect based approaches) would be computationally challenging; this challenge is compounded by the data being stored in a trusted research environment limiting the computational power and tools available.
Second, the data is coarse with most episodes having only a single positive test; therefore, the trajectory of the viral load would not be well-identified in an individual.
Combined, these factors outweighed the downside of survival analysis only considering the binary outcome (\ie positive or negative) or the test rather than the additional information on the viral load.

The methods in this chapter deal with modelling two elements of the data-generating process, double interval censoring and truncation.
False negatives are not dealt with and left to \cref{E-imperf-test}.

@@

Truncation is a large problem within CIS, especially when individuals move to the more infrequent testing schedule (described in \cref{E-intro:sec:cis}).
Consider that, based on the estimates in \cref{E-ATACCC}, the median length of an episode is 14 days \todo{check this number} we would not expect many episodes to last the 28 days that can occur between tests.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/truncation}
  \caption{Episodes data in CIS is truncated, meaning that an unknown number of infections are never detected. Consider the two individuals shown here: we collect identical data for them (a series of negative tests) yet the top individual was never infected and the bottom individual was. See the main text for why this is an issue. \label{perf-test:fig:truncation}}
\end{figure}

@@

While this chapter sets up the framework for analysing the CIS data, applying the method to CIS data requires modelling false negatives.
That is the subject of \cref{E-imperf-test} and hence a full description of the data and processing of it is deferred to that chapter.

\section{Modelling the duration}\label{perf-test:sec:model}
In this section, I develop a model accounting for double interval censoring and truncation, building on prior work in both biomedical and bird nesting studies.
The arbitrary nature of truncation in CIS, where each individual has a unique truncation pattern and is neither simple left nor right truncation, has not previously been explored in the literature.
Furthermore, I explore a Bayesian framework where the amount of explicit data augmentation is limited allowing the method to scale to the large number of episodes present in CIS without requiring vast computing resources (such as a high-performance cluster), as only limited resources are available within the secure research service (SRS, a form of trusted research environment) where the CIS data is held.
Therefore, unlike most Bayesian methods (e.g.: \textcite{heBayesiana}, \textcite{heBayesian}, and \textcite{caoModeling}) we avoid augmenting the data with unobserved times of infection or any information on the truncated infections; this would be computationally expensive with the large number of episodes in CIS (>4000 in the time period considered in the main analysis of \cref{E-imperf-test}).

@@

I have also considered a model including individuals without any observed episodes (see \cref{E-appendix:total-model} for details).
These individuals may have truncated episodes or may not have been infected at all.
However, I found that this model has little difference in results while increasing computational complexity and requiring assumptions on reinfection (discussed in \cref{perf-test:sec:discussion}).

@@

In this section, I consider the various parameterisation of the survival function that are available.
This has implications for the choice of prior it is not possible to be vague on all sensible parameterisations.
Informative priors might also be attractive, and allow the incorporation of the estimates from \cref{E-ATACCC}.

@@

An uninformative prior would be $\lambda_t \dist \Beta(\alpha, \alpha)$, with common choices of $\alpha$ being $0.5$, Jeffreys' Prior, or $1$, a flat prior.
Flat or uninformative priors on the hazard lead to very informative priors on the survival, favouring short survival times (see \cref{perf-test:fig:flat-prior}).
The sharp falloff can be seen by noting that these priors have expectation 0.5, hence $\E(S(t)) = \E \left( \prod_{i=1}^{t-1} (1-\lambda_t) \right) = 0.5^{t-1}$ which decreases very quickly.
This prior expresses extreme scepticism in estimates from previous studies, such as \cref{E-ATACCC} and the meta-analysis of \textcite{cevikShedding}.
Therefore, I favour a prior which is weakly informative on the survival, or an informative prior used to represent our belief from the \cref{E-ATACCC} analysis.
\begin{figure}
  \centering \includegraphics{cis-perfect-testing/flat-prior}
  \caption{The density of Beta(1, 1) flat prior and Beta(0.5, 0.5) uninformative priors (left) with the prior predictive survival time (right). The implied prior on the survival time is very short when using either prior on the hazard at each time. \label{perf-test:fig:flat-prior}}
\end{figure}

@@

The analysis in \cref{E-ATACCC} can be used to provide prior information for the CIS analysis.
In particular, the ATACCC study frequently samples individuals early in the infection.
Therefore, the survival distribution from that analysis should be reliable through this period.
CIS can then inform the remainder of the distribution where the previous analysis is based on extrapolation.

The model structure assumed in \cref{E-ATACCC} leads to positive correlation in the posterior estimates of the hazard.
It is important to propagate this correlation as part of the prior uncertainty into the present analysis.
We do this in two parts: first, approximate the previous posterior estimate of the logit-hazard as a multivariate normal (which can be viewed as an approximation of Markov melding~\autocite{goudieJoining}).
Then, use a Beta process to add additional uncertainty to this.

There are two reasons to increase the uncertainty relative to the \cref{E-ATACCC} posterior.
First, we do not have full belief in these estimates due to the extrapolation problem previously discussed.
Second, the results may not fully generalise between the two studies.
The samples are tested using different laboratories, with different PCR machines, and hence may not have the same characteristics~\autocite{punchooLaboratory}.
The study participants, while both targeting a general population cohort, may also differ because ATACCC recruited through NHS Test and Trace while CIS recruited by sampling from a database of addresses (see \cref{E-intro:sec:data}).

The first step, approximating the \cref{E-ATACCC} posterior hazard as a multivariate normal on the logit scale, is done through the method of moments.
The approximation is very good (see \cref{perf-test:fig:approximate-ATACCC}).

@@

I take as the ground truth distribution for duration of positivity a combination of the estimates from \cref{E-ATACCC} with an inflated tail to represent what is seen within CIS.
The tail is modified based on Sarah Walker's (unpublished) work based on CIS data.
These curves and the combined curve are compared in \cref{perf-test:fig:duration-dist}.
Individual $i$'s duration of positivity, $D_i$, is then an independent draw from this distribution.

@@

\ifSubfilesClassLoaded{
  \appendix
  \subfile{appendix-cis-perfect-testing}
  \listoftodos
}{}
\end{document}
